---
title: "Pipeline"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in data and create `SummarizedExperiment` objects

A SummarizedExperiment object will serve as input for the App. There most be 
a slot for raw "counts" and a grouping factor named "group" in colData. Other
colData variables are allowed but "group" must be present.

```{r}
library(here)
library(tidyverse)


hh1_counts <- readRDS(here("data", "hh1-gene-mat.rds"))
hh1_annotation <- read.csv(here("doc", "hh1-annotation.csv"))
hh1_col_data <- hh1_annotation[, c("treatment", "time_hr")]
hh1_col_data$time_hr <- factor(as.numeric(hh1_col_data$time_hr))
rownames(hh1_col_data) <- hh1_annotation$sample_name

# match column ordering
hh1_counts <- hh1_counts[, rownames(hh1_col_data)]
```

## Create SummarizedExperiment class (use HH1 experiment as example)

The App will only accept `SummarizedExperiment` data types!

```{r}
library(SummarizedExperiment)


raw <- SummarizedExperiment(assays = list(counts = hh1_counts), colData = hh1_col_data)
raw$group <- factor(str_c(raw$treatment, raw$time_hr, sep = "."))

# save for testing
saveRDS(raw, here("data", "raw.rds"))
```

## Filtering

For now, allow filtering through the `filterByExpr` function. 

```{r}
# simulate user input
group_data <- raw$group
min_count <- 10
min_total_count <- 15
min_prop <- 0.7

# view dimension before filtering
dim(raw)

# filter the counts
keep <- edgeR::filterByExpr(raw, 
                            group = group_data,
                            min.count = min_count,
                            min.total.count = min_total_count,
                            min.prop = min_prop)

filt <- raw[keep, ]
dim(filt)
```

## Visualize distributions before and after

```{r}
# simulate user selecting a sample
user_sample <- colnames(raw)[[1]]

# Create plotting function
plot_distribution <- function(se, sample_name, n_bins = 30, pseudocount = 0) {
  assays(se)[["counts"]][, user_sample] %>% 
    enframe(name = NULL, value = "x") %>% 
    mutate(x = x + pseudocount) %>% 
    ggplot(aes(x)) +
    geom_histogram(bins = n_bins, fill = "steelblue", color = "black") +
    theme_light() +
    scale_x_log10() +
    labs(x = "Number of Reads")
}

# example showing before/after filtering
plot_distribution(raw, user_sample) + ggtitle(paste(user_sample, "Distribution of Raw Counts"))
plot_distribution(filt, user_sample) + ggtitle(paste(user_sample, "Distribution of Filtered Counts"))

# maybe show summary stats too -------------------------------------------------
dist_summary <- function(se, sample_name, pseudocount = 0) {
  summary(assays(se)[["counts"]][, user_sample] + pseudocount)
}

dist_summary(raw, user_sample)
dist_summary(filt, user_sample)
```

## Normalize the filtered dataset

```{r}
# user defined pseudocount for log2 transformation
pseudocount <- 2

# Total count ------------------------------------------------------------------
tc <- edgeR::cpm(assays(filt)[["counts"]])
tc_lcpm <- edgeR::cpm(assays(filt)[["counts"]], log = TRUE, prior.count = pseudocount)

assays(filt)[["TC"]] <- tc
assays(filt)[["logTC"]] <- tc_lcpm

# edgeR ------------------------------------------------------------------------
percentile <- 0.7
reference_col <- NULL

tmm <- edgeR::calcNormFactors(filt, method = "TMM", refColumn = reference_col)
uq <- edgeR::calcNormFactors(filt, method = "upperquartile", p = percentile)
rle <- edgeR::calcNormFactors(filt, method = "RLE")

tmm_cpm <- edgeR::cpm(tmm)
uq_cpm <- edgeR::cpm(uq)
rle_cpm <- edgeR::cpm(rle)
tmm_lcpm <- edgeR::cpm(tmm, log = TRUE, prior.count = pseudocount)
uq_lcpm <- edgeR::cpm(uq, log = TRUE, prior.count = pseudocount)
rle_lcpm <- edgeR::cpm(rle, log = TRUE, prior.count = pseudocount)

assays(filt)[["TMM"]] <- tmm_cpm
assays(filt)[["UQ"]] <- uq_cpm
assays(filt)[["RLE"]] <- rle_cpm
assays(filt)[["logTMM"]] <- tmm_lcpm
assays(filt)[["logUQ"]] <- uq_lcpm
assays(filt)[["logRLE"]] <- rle_lcpm

# QSmooth ----------------------------------------------------------------------
qs <- qsmooth::qsmooth(filt, group_factor = filt$group)
qs_cpm <- edgeR::cpm(qsmooth::qsmoothData(qs))
qs_lcpm <- edgeR::cpm(qsmooth::qsmoothData(qs), log = TRUE, prior.count = pseudocount)

assays(filt)[["QS"]] <- qs_cpm
assays(filt)[["logQS"]] <- qs_lcpm

# RUVg -------------------------------------------------------------------------
# This would be a user supplied vector
control_genes <- c("ACTB", "B2M", "C1orf43", "CHMP2A", "GAPDH", "GPI", "GUSB", 
                   "HPRT1", "HSP90AB1", "PPIA", "PSMB2", "PSMB4", "RAB7A", "REEP5", 
                   "RPL13A", "RPLP0", "SNRPD3", "TFRC", "UBC", "VCP", "VPS29")
K = 3

ruv_set <- RUVSeq::RUVg(as.matrix(assays(filt)[["counts"]]), control_genes, k = K)
ruv_cpm <- edgeR::cpm(ruv_set$normalizedCounts)
ruv_lcpm <- edgeR::cpm(ruv_set$normalizedCounts, log = TRUE, prior.count = pseudocount)

assays(filt)[["RUVg"]] <- ruv_cpm
assays(filt)[["logRUVg"]] <- ruv_lcpm
```

## Create boxplots of the normalized counts

```{r}
plot_rle <- function(assay_name, fill_by = "group", outlier_shape = NA, outlier_alpha = 0.5) {
  median_expr <- apply(assays(filt)[[assay_name]], 1, median)
  rle_data <- assays(filt)[[assay_name]] - median_expr
  
  plot_df <- rle_data %>% 
    dplyr::as_tibble(rownames = "feature_id") %>% 
    pivot_longer(-feature_id, names_to = "sample_name", values_to = "RLE") %>% 
    left_join(as_tibble(colData(filt), rownames = "sample_name"), 
              by = "sample_name")
  
  plot_df %>% 
    ggplot(aes(x = sample_name, y = RLE, fill = .data[[fill_by]])) +
    geom_boxplot(outlier.shape = outlier_shape, outlier.alpha = outlier_alpha) +
    theme_light() +
    geom_hline(yintercept = median(plot_df$RLE, na.rm = TRUE), 
               linetype = 2,
               color = "red") +
    labs(title = paste(assay_name),
         x = NULL,
         y = "RLE") +
    theme(axis.text.x = element_blank())
}

# ------------------------------------------------------------------------------
plot_rle("logTMM", "treatment", outlier_shape = 19)
plot_rle("logRLE", "treatment", outlier_shape = 19)
plot_rle("logUQ", "treatment", outlier_shape = 19)
plot_rle("logQS", "treatment", outlier_shape = 19)
plot_rle("logRUVg", "treatment", outlier_shape = 19)
plot_rle("logTC", "treatment", outlier_shape = 19)
```

## Sample vs Sample Scatter Plots

```{r}
# simulate user sample selections
sample_1 <- "dmso.96.3"
sample_2 <- "dmso.96.1"

# ------------------------------------------------------------------------------
plot_scatter <- function(assay_name, sample1, sample2, pt_alpha = 0.5, log_scale = TRUE) {
  plot_df <- assays(filt)[[assay_name]] %>% 
    as_tibble(rownames = "feature_id") %>% 
    select(feature_id, sample1, sample2)
  
  if (log_scale) {
    plot_df %>%
      ggplot(aes(x = .data[[sample1]], y = .data[[sample2]])) +
      geom_point(alpha = pt_alpha) +
      geom_abline(slope = 1, intercept = 0, color = "red", linetype = 2) +
      scale_x_log10() +
      scale_y_log10() +
      theme_light() +
      labs(title = paste(assay_name))
  } else {
    plot_df %>%
      ggplot(aes(x = .data[[sample1]], y = .data[[sample2]])) +
      geom_point(alpha = pt_alpha) +
      geom_abline(slope = 1, intercept = 0, color = "red", linetype = 2) +
      theme_light() +
      labs(title = paste(assay_name))
  }
}

# test -------------------------------------------------------------------------
plot_scatter("TMM", sample_1, sample_2, log_scale = TRUE, pt_alpha = 0.05)
plot_scatter("RLE", sample_1, sample_2, log_scale = TRUE, pt_alpha = 0.5)
plot_scatter("UQ", sample_1, sample_2, log_scale = TRUE, pt_alpha = 0.5)
plot_scatter("QS", sample_1, sample_2, log_scale = TRUE, pt_alpha = 0.5)
plot_scatter("RUVg", sample_1, sample_2, log_scale = TRUE, pt_alpha = 0.5)
plot_scatter("TC", sample_1, sample_2, log_scale = TRUE, pt_alpha = 0.5)
```

## Smear Plots

```{r}
plot_smear <- function(assay_name, sample1, sample2, smooth = FALSE, loess = FALSE) {
  X <- assays(filt)[[assay_name]][, c(sample1)]
  Y <- assays(filt)[[assay_name]][, c(sample2)]
  edgeR::maPlot(X, 
                Y, 
                normalize = FALSE,
                smooth.scatter = smooth,
                lowess = loess,
                main = paste(assay_name, "-", sample1, "vs.", sample2), 
                ylab = paste0("log2(", sample2, " / ", sample1, ")"))
  abline(a = 0, b = 0, lty = 2)
}

# ------------------------------------------------------------------------------
plot_smear("TMM", sample_1, sample_2, loess = TRUE)
plot_smear("RLE", sample_1, sample_2, loess = TRUE)
plot_smear("UQ", sample_1, sample_2, loess = TRUE)
plot_smear("QS", sample_1, sample_2, loess = TRUE)
plot_smear("RUVg", sample_1, sample_2, loess = TRUE)
plot_smear("TC", sample_1, sample_2, loess = TRUE)
```

## PCA plots

```{r}
plot_pca <- function(assay_name, 
                     scale_data = TRUE, 
                     center_data = TRUE, 
                     component1 = "PC1", 
                     component2 = "PC2", 
                     color_by = "group",
                     shape_by = NULL) {
  pca <- prcomp(t(assays(filt)[[assay_name]]), center = center_data, scale. = scale_data)
  pca_df <- as_tibble(pca$x, rownames = "sample_name") %>% 
    left_join(as_tibble(colData(filt), rownames = "sample_name"), by = "sample_name")
  
  if (is.null(shape_by)) {
    pca_df %>% 
      ggplot(aes(x = .data[[component1]], y = .data[[component2]], color = .data[[color_by]])) +
      geom_point(size = 3) +
      geom_hline(yintercept = 0, linetype = 2) +
      geom_vline(xintercept = 0, linetype = 2) +
      theme_light() +
      labs(title = paste(assay_name))
  } else {
    pca_df %>% 
      ggplot(aes(x = .data[[component1]], y = .data[[component2]], color = .data[[color_by]], shape = .data[[shape_by]])) +
      geom_point(size = 3) +
      geom_hline(yintercept = 0, linetype = 2) +
      geom_vline(xintercept = 0, linetype = 2) +
      theme_light() +
      labs(title = paste(assay_name))
  }
}

# ------------------------------------------------------------------------------
plot_pca("logTMM", color_by = "treatment", shape_by = "time_hr")
plot_pca("logRLE", color_by = "treatment", shape_by = "time_hr")
plot_pca("logUQ", color_by = "treatment", shape_by = "time_hr")
plot_pca("logQS", color_by = "treatment", shape_by = "time_hr")
plot_pca("logRUVg", color_by = "treatment", shape_by = "time_hr")
plot_pca("logTC", color_by = "treatment", shape_by = "time_hr")
```

## Correlation plots

```{r}
library(corrplot)


plot_cor <- function(se, assay_name, samples, viz_method) {
  cor_data <- cor(assays(se)[[assay_name]][, samples])
  corrplot::corrplot(
    cor_data,
    method = viz_method,
    type = "lower", 
    col = viridis::viridis(100), 
    outline = TRUE,
    tl.col = "black",
    tl.srt = 45,
    diag = FALSE)
}

plot_cor(filt, "TMM", colnames(filt)[c(2, 4, 6, 8, 10)], "circle")
```

## Heatmap on most variable genes

```{r}
plot_heatmap <- function(se, assay_name, samples, n_features, cd_rows, cd_cols, cd_method) {
  mat <- assays(se)[[assay_name]]
  row_vars <- matrixStats::rowVars(mat)
  var_order <- order(row_vars, decreasing = TRUE)
  var_mat <- mat[head(var_order, n_features), samples]
  pheatmap::pheatmap(var_mat, 
                     scale = "row",
                     color = colorRampPalette(c("dodgerblue3", "grey99", "firebrick3"))(50),
                     show_rownames = FALSE,
                     clustering_distance_rows = cd_rows,
                     clustering_distance_cols = cd_cols,
                     clustering_method = cd_method)
}
  
plot_heatmap(filt, "logTMM", colnames(filt)[1:10], 1000, "correlation", "correlation", "ward.D2")  
```

